---
title: "Day 2 - Practical 2 - more advanced methods for metabolomics data analysis"
subtitle: "2024 NUTRIM microbiome & metabolome workshop"
author: YOUR-NAME
date: last-modified
keep-md: false
embed-resources: true
format: html
theme: 
  light: flatly
  dark: darkly
toc: true
number-sections: true
---

## 1A. Intro + using Quarto {#using-quarto}

This is a notebook for the second tutorial on metabolomics! Feel free to add notes and text where you want and code chunks to execute and copy the code.

Please open .html file with the instructions for this tutorial , that you can find here: TODO

Remember to use the Visual mode of the notebook and uncheck "Render on Save" checkbox above!

## 1B. Using Quarto - reminder2 📝

During this workshop, we will write code and notes in Quarto documents.

Write notes in "[markdown](https://quarto.org/docs/authoring/markdown-basics.html "Markdown is plain text, with special characters to indicate headings, bold text, links etc"){target="_blank"}", e.g. your intentions, hypotheses, observations, etc.

```{r}
# write code in "chunks" (and comments too, using the # symbol)
print("Hello World")
```

Insert a chunk with option +cmd + I (Mac) or ctrl+ shift + I (Windows)

```{r, eval=FALSE}
print("Run this one line with ctrl + enter, or cmd + enter on macOS")
print("or run a whole chunk with ctrl/cmd + SHIFT + enter")
```

Run all previous chunks by clicking the first button in the corner of the chunk.

You can add notes wherever you want and new code chunks!

## 2. Load R packages and source functions 📦

```{r}
library(here)
library(tidyverse) # data cleaning 
library(ggplot2) # visualizations
source(here("functions.R"))  # URF functions
library(randomForest) 
library(factoextra) 
library(limpca) # ASCA
```

## 3. Read the data 📂

```{r}
# Read metabolites data table
metabolites <- read.csv(file = here("data:papa2012/processed/metabolites_short.csv"), row.names = 1)

# PQN normalize it
metabolites_pqn <- pqn(metabolites)

# Read metadata
metadata_metabolites <- read.csv(file = here("data:papa2012/processed/metadata_metabolites.csv"), row.names = 1)
```

## 4. URF 🌲🌲🌲

**TASK 1:**

**TASK 2:**

Perform URF:

```{r}

```

**TASK 3:**

Look at the structure of the resulting URF model:

```{r}
str(URF)
```

**TASK 4:**

Plot URF score plot:

```{r}

```

## 5. ASCA

**TASK 1:**

Read metadata_metabolites_asca.csv. How many samples do we have now?

```{r}
# plot the URF score plot

```

**TASK 2:**

```{r}
metabolites_asca <- metabolites_pqn[which(rownames(metabolites_pqn) %in% metadata_metabolites_asca$SampleID), ]
```

**TASK 3:**

Create `Age_group` vector:

```{r}

```

**TASK 3:**

Check the data balance:

```{r}

```

### 5.2 Create an ASCA object

Create an empty list:

```{r}
ASCA_model <- list()
```

Add outcomes:

```{r}
ASCA_model <- list()
```

Add design:

```{r}
ASCA_model$design <- metadata_metabolites_asca %>% 
  dplyr::select(c("Diagnosis", "Age_group")) %>%
  dplyr::mutate(Diagnosis = factor(Diagnosis, levels = c("nonIBD", "CD", "UC")),
                Age_group = factor(Age_group, levels = c("minor", "adult")))

# Row names of the design matrix need to be the same as of the outcomes (sample names)
rownames(ASCA_model$design) <- metadata_metabolites_asca$SampleID
```

Plot design matrix:

```{r}
plotDesign(design = ASCA_model$design, x = "Diagnosis", y = "Age_group",
    title = "Design of the metabolomics dataset")
```

Add formula:

```{r}
ASCA_model$formula <- "outcomes ~ Diagnosis + Age_group + Diagnosis:Age_group"
```

### 5.3 Perform ASCA

**5.3.1 Generate and print the model matrix:**

```{r}
resLmpModelMatrix <- lmpModelMatrix(ASCA_model)
head(resLmpModelMatrix$modelMatrix)
```

**5.3.2 Decompose the model matrix:**

```{r}
resLmpEffectMatrices <- lmpEffectMatrices(resLmpModelMatrix)
```

**5.3.3 ASCA**

```{r}
resASCA <- lmpPcaEffects(resLmpEffectMatrices = resLmpEffectMatrices,
                         method = "ASCA")
```

**5.3.4 Check contributions:**

```{r}
resLmpContributions <- lmpContributions(resASCA)
```

-   Total variance

```{r}
resLmpContributions <- lmpContributions(resASCA)

resLmpContributions$totalContribTable
```

-   PCs contributions in each model

```{r}
resLmpContributions$effectTable
```

**5.3.5 Statistical testing**

```{r}
resLmpBootstrapTests <- lmpBootstrapTests(resLmpEffectMatrices = resLmpEffectMatrices, 
                                          nboot = 100)

t(resLmpBootstrapTests$resultsTable)
```

\*\*5.3.6 Scores and loadings plots\*

```{r}
lmpScorePlot(resASCA, effectNames = "Diagnosis",
                          color = "Diagnosis")

lmpLoading1dPlot(resASCA, effectNames = "Diagnosis",
                 axes = 1, # only PC1
                  xaxis_type = "character",)
```

## 6. Finish! ⏩

-   Good job! 🎉 You have finished this tutorial! Now you can use also URF and ASCA in R!

## 7. Session info

```{r}
sessioninfo::session_info()
```
