---
title: "Day 2 - Practical 2 - more advanced methods for metabolomics data analysis"
subtitle: "2024 NUTRIM microbiome & metabolome workshop"
author: Michal Skawinski
editor: visual
date: last-modified
keep-md: false
theme: 
  light: flatly
  dark: darkly
format: html
embed-resources: true
code-block-border-left: true
code-block-bg: true
toc: true
toc-location: right
toc-depth: 4
toc-expand: 3
number-sections: true
number-depth: 3
fig-align: center
fig-dpi: 300
fig-width: 7.5
fig-height: 5
fig-responsive: true
code-tools: true
code-fold: false
code-link: true
code-annotations: below
lightbox: auto
cache: false
bibliography: references.bib
---

# Intro 🔆

In this tutorial, we are going to explore some more advanced methods and how to implement them in R:

-   [**URF**](#urf) (Unsupervised Random Forest), and

-   [**ASCA**](#asca) (ANalysis Of VAriance (ANOVA)-Simultaneous Component Analysis).

# Load R packages and some functions 📦

```{r}
#| output: false
library(here)
library(tidyverse) # data cleaning 
library(ggplot2) # visualizations
source(here("functions.R"))  # URF functions
library(randomForest) 
library(factoextra) 
library(limpca) # ASCA

```

# Read the data 🔍

We are going to use the same data as before, so you should have it already stored in your environment. However if you restarted the R session, you can use the code below to read the data again:

```{r}
# Read metabolites data table
metabolites <- read.csv(file = here("data:papa2012/processed/metabolites_short.csv"), row.names = 1)

# PQN normalize it
metabolites_pqn <- pqn(metabolites)

# Read metadata
metadata_metabolites <- read.csv(file = here("data:papa2012/processed/metadata_metabolites.csv"), row.names = 1)
```

# URF 🌲🌲🌲 {#urf}

We have provided a function to perform URF in R:

```{r}
#| output: asis
#| eval: false
script_URF(nr_itteration, real_data, nr_trees, nr_samples, class) 
```

##### **TASK 1**

View the code of the function. You don't need to understand or analyze the body of the function. Just look at the comments and try to analyze the inputs and outputs:

```{r}
print(script_URF)
```

What are the input arguments ? What are the outcome(s) of the function? Tip: Look what's inside the `return`.

##### **TASK 2**

Let's perform our first URF! Build a function with 1500 trees and 8 samples in a terminal node. Limit the number of iterations up to 10, which will speed up the computations (generally we would use around 100 iterations):

::: {.callout-note collapse="true"}
### Do you need to scale the data?

URF is an non-linear method, therefore it will perform well with non-scaled data.
:::

```{r}
# write your code here
```

::: {.callout-note collapse="true" appearance="simple"}
##### SOLUTION:

```{r}
#| message: false
#| output: false
URF <- script_URF(nr_itteration = 1, real_data = metabolites_pqn, nr_trees = 1500, nr_samples = 8)
```
:::

##### **TASK 3**

Look at the structure of the resulting URF model:

```{r}
str(URF)
```

##### **TASK 4**

Plot the URF score plot. Color the points by `Diagnosis`.

```{r}
ggplot(as.data.frame(URF$pc$x), 
       aes(x = PC1, 
           y = PC2,
           color = metadata_metabolites$Diagnosis)) + 
  geom_point() + 
  labs(x = paste0("PC1 (",round(URF$pr[1,2],2),"%)"), 
       y = paste0("PC2 (",round(URF$pr[2,2],2),"%)")) + 
  theme_bw()
```



# ASCA {#asca}

There are many versions of ASCA, depending on the application and **data design**. Today, we will only familiarize with one version: ASCA+ from the R package `limpca` [@thiel2023].

More information about the package, as well as tutorials and examples with datasets can be found [here](https://bioconductor.org/packages/release/bioc/html/limpca.html).

## ASCA - case study

We are going to inspect the effect of the disease (nonIBD, UC, CD) and the duration of the disease on the metabolites' profiles. 

In our data, we have multiple samples per participant. This comes from sampling of patients at different times. However, we are not interested in in as it doesn't follow any design.

Therefore, we have provided a new file `metadata_metabolites_asca.csv` with the metadata of only selected samples from the same time points, which contains an additional variable, `Age_at_diagnosis`, that we are interested in here.

##### **TASK 1**

Read `metadata_metabolites_asca.csv`. How many samples do we have now?
```{r}
metadata_metabolites_asca <- read.csv(file = here("data:papa2012/asca/metadata_metabolites_asca.csv"), row.names = 1)
nrow(metadata_metabolites_asca)
```


##### **TASK 2**

We also need to select those samples from the metabolites data table (PQN normalized).
```{r}
metabolites_asca <- metabolites_pqn[which(rownames(metabolites_pqn) %in% metadata_metabolites_asca$SampleID), ]
```

##### **TASK 3**

In the metadata, we don't have information on the duration of the disease. However, we know participants age and the age at diagnosis. Calculate the duration of the disease and assign it to a new vector `duration_of_disease`. Then, add the vector to `metadata_metabolites_asca`.
```{r}
# write your code below

```

::: {.callout-note collapse="true" appearance="simple"}
##### SOLUTION:

```{r}
duration_of_disease <- metadata_metabolites_asca$Age - metadata_metabolites_asca$Age_at_diagnosis

metadata_metabolites_asca$duration_of_disease <- duration_of_disease
```
:::



## Create an ASCA object

In order to use ASCA from the `limpca` package, the data need to be formatted as a list with the following elements:

-   outcomes (multivariate matrix),

-   design (data.frame) and

-   formula (character string).

First, create an empty list:
```{r}
ASCA_model <- list()
```

##### **Outcomes**
TODO 


##### **Design**
TODO
```{r}
ASCA_model$design <- data.frame(metadata_metabolites)
```

##### **Formula**

TODO

## Perform ASCA




table(metadata_metabolites_small$diagnosis[(which(metadata_metabolites_small$consent_age \<18))\])

table(metadata_metabolites_small$diagnosis[(which(metadata_metabolites_small$consent_age \>= 18 & metadata_metabolites_small\$consent_age \<= 60))\])



# Session info

<details>

```{r}
sessioninfo::session_info()
```

</details>
